Fail Faster

### 2025-03-22 paper/paper.tex
bleh
some wibbles on the intro

This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that a generator uses to make choices can
constitute an unexpectedly high proportion of its runtime. These factors combine to limit 
the effectiveness of PBT.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi

We chart a path forward.
First, we show how to use \textit{multi-stage programming}, a classic technique for 
building fast DSLs “without regret”~\cn{}, to eliminate the overhead of 
many common generator abstractions. We refer to this approach as \name.
Second, we demonstrate that a generator's source of entropy is an underappreciated design factor by 
isolating its runtime cost and showing that significant speedups can be achieved 
while preserving output equivalence.
These insights lay the groundwork for generator libraries that combine high-level 
design with low-level performance.

@@

To evaluate our technique, we implement staged PBT libraries in two strict functional languages: OCaml and Scala 3.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library, and by far the most widely-used\cn{}.
In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using ScalaCheck or \bq, respectively.
This semantic equivalence enables pointwise, as opposed to distributional, comparisons, ensuring that performance differences are attributable \textit{solely} to staging and \rand optimizations.
% \jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that. ``Comparing distributions of speed is hard, comparing pointwise is easier.''. This needs to have its own sentence here.}

We evaluate the optimizations with a series of case studies in each
generator library, and show that the \name generators run faster than their unstaged equivalents.
Further, we show generators that run faster also find bugs faster
% ability \jwc{The paper is called ``fail faster'' --- we shouldn't say ``bug finding ability'', we should say something like ``how fast you can squash bugs''. This is too passive }
by running these case studies in the Etna
platform~\cite{etna}, which simulates real-world PBT usage \iflater\hg{Don't love this phrase, I'll
think more about how we should talk about Etna}\fi by injecting bugs into a system and
measuring how quickly various generators detect them.

>>>>

This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that a generator uses to make choices can
constitute an unexpectedly high proportion of its runtime. These factors combine to limit 
the effectiveness of PBT. \jwc{Unsure if ``constitute a high proportion of its runtime'' limits effectiveness. it's more that the choice \rand limits effectiveness.}
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi

We study and address these issues separately.
First, we show how to use \textit{multi-stage programming}, a classic technique for 
building fast DSLs “without regret”~\cn{}, to eliminate the overhead of 
many common generator abstractions. We refer to this approach as \name.
% Secondly, we demonstrate the
% impact of diferent choices of randomness library on generator performance by conducting a
% controlled comparison.
Second, we demonstrate that a DSL's choice of \rand{} has a
dramatic impact on its performance by performing a controlled experiment,
isolating the performance benefits of faster randomness in a clearly quantifiable manner.
\jwc{Another cut at this.}

% These insights lay the groundwork for generator libraries that combine high-level 
% design with low-level performance.

@@

To demonstrate the multi-language applicability of \name{}, we implement staged generator DSLs in two strict functional languages: OCaml and Scala 3.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library\cn{}.
In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using ScalaCheck or \bq, respectively.
This semantic equivalence enables pointwise---as opposed to distributional---bug-finding comparisons, ensuring that if a \name{} generator finds bugs faster, this is attributable \textit{solely} to staging and \rand optimizations, not luck with seeds.
% \jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that. ``Comparing distributions of speed is hard, comparing pointwise is easier.''. This needs to have its own sentence here.}

We evaluate the optimizations with a series of case studies in each
generator library, and show that the \name{} generators run faster than their unstaged equivalents.
Further, we show generators that run faster also find bugs faster
% ability \jwc{The paper is called ``fail faster'' --- we shouldn't say ``bug finding ability'', we should say something like ``how fast you can squash bugs''. This is too passive }
by running these case studies in the Etna
platform~\cite{etna}, which simulates real-world PBT usage \iflater\hg{Don't love this phrase, I'll
think more about how we should talk about Etna}\fi by injecting bugs into a system and
measuring how quickly various generators detect them.
